\documentclass{llncs}

% name of the language
\newcommand{\lang}[0]{Mediator}
\newcommand{\emlang}[0]{\emph{\lang{}}}

\title{Component-Based Modeling in \lang{}}

\author{Yi Li\and Meng Sun}
\institute{LMAM and Department of Informatics, School of Mathematical Sciences, \\ Peking University, Beijing, China \\
\email{liyi\_math@pku.edu.cn, sunmeng@math.pku.edu.cn}
}


\input{common}

\begin{document}

\maketitle

\begin{abstract}
In this paper we propose a new language \emph{\lang{}} to formalize component-based system models. \lang{} supports a two-step modeling approach. \emph{Automata}, encapsulated with an interface of ports, are the basic behavior units. \emph{Systems} declare components or connectors through automata, and glue them together. With the help of \lang{}, components and systems can be modeled separately and precisely. Through various examples, we show that this language can be used in practical scenarios.

\keywords{Component-based Modeling, Coordination, Formal Method}
\end{abstract}

\input{sections/introduction}
% \input{sections/overview}
\input{sections/grammar}
\input{sections/semantics}
\input{sections/casestudy}

\section{Conclusion and Future Work}
\label{sec:conclusion}

A new modeling language \lang{} is proposed in this paper to help with component-based software engineering through a formal way. With the basic behavior unit \emph{automata} that captures the formal nature of components and connections, and \emph{systems} for hierarchical composition, the language is easy-to-use for both formal method researchers and system designers.

This paper is a preface of a set of under-development tools. We plan to build a model checker for \lang{}, and extend it through symbolic approach. An automatic code-generator is also being built to generate platform-specific codes like \emph{Arduino}\cite{margolis2011arduino}.

\section*{Acknowledgements}

The work was partially supported by the National Natural Science Foundation of China under grant no. 61532019, 61202069 and 61272160.

\bibliographystyle{splncs03}
\bibliography{fm,local}

\newpage
\input{sections/appendix}

\end{document}